The Role of Advection Versus Diffusion in Intraventricular Transport

### 2025-02-25 paper/main.tex
update figures, add force calculation

@@

The CSF velocities are most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity magnitude drops from 27.7 \textmu m/s to 8.0 \textmu m/s.
In vivo, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic cilia has minimal
impact and results in maximum speeds of 27.7 \textmu m/s or 27.2 \textmu m/s, respectively.
Of the total tangential force exerted by the cilia, ventral cilia in the middle ventricle
contribute with 34\%, while the dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). The magnitude of the force contributions are thus
more similar than the difference in their impact on the maximum velocity magnitudes.

Changes in the transport dynamics of Dendra2 are most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit changes in the dynamics of $\cbar$ (\Crefpanels{fig:fig5}{c}).
Compared to the baseline model results, we observe a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
In contrast, removal of telencephalic cilia does not alter the evolution of the
mean concentration $\cbar$ with time (\Crefpanels{fig:fig5}{c}).
Finally, without ventral cilia in the diencephalic ventricle,
more solute spreads towards the anterior ventricle (ROI 5) and
less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} The total tangential force applied through the tangential traction
    boundary condition in the three separate ciliated regions.
    \textbf{c.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

\subsection{Modification of ventricles geometry impacts protein distribution}
Variations in ventricular morphology impact CSF flow patterns.
To assess how ventricular morphologies influence solute distribution,
we consider four alternative brain ventricles meshes:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Crefpanels{fig:fig6}{a}).
We simulate CSF flow in the modified geometries with the baseline flow model
and use the resulting velocity field to simulate transport of Dendra2
(\Crefpanels{fig:fig6}{b}).

Constricting the connection between the anterior and middle ventricle leads to
slower transport to the anterior ventricle (ROI 5),
and a slightly higher mean concentration in ROIs 2--4 and 6, compared to the original geometry.
On the other hand, constricting the duct connecting the middle and the posterior ventricle
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a shrunk middle ventricle.
Additionally, for the latter configuration, the transport upstream of CSF flow towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.
For the mesh where all three ventricles are modified,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while both the anterior and the posterior ventricles (ROIs 5 and 6) have lower mean concentrations.
Altogether, our results show that the geometry and width of the ventricular ducts matter
for solute transport. This may explain why we observed larger differences
in the cilia-deficient computational model compared to the experimental data in
\Cref{fig:fig4}, as the model and the fish have different duct diameters.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{\textbf{a.} The original computational mesh of the ventricular geometry.
    \textbf{b.} The four modified versions of the computational geometry. Compared
    to the original mesh, the volumes of the deformed meshes are reduced by 3.9\%, 4.1\%, 1.4\% 
    and 9.4\%.
    \textbf{c.} Mean concentration $\cbar(t)$ simulated with the original and modified geometries
    in the 6 ROIs. 
    \mk{MK: How do the different geoemtries compare in terms of volume,
    areas where cilia are active?} \lyng{HH: Good question, will add this information.}}
    \label{fig:fig6}
\end{figure}

@@

\subsection{Limitations and further work}
The model presented has certain limitations. Future work could consider:
\begin{itemize}
    \item spatial heterogeneity of the cilia, how this affects flow and transport
    \item time-varying cilia?
    \item improved advection-diffusion equation influx/outflux BCs, or instead
    modeling deformation of a closed ventricular system. 
    and investigating how the arising flow compares with pulsatile flow observed in experiments
    \item FSI: CSF flow coupled to 1D/2D/3D cilia structures.
\end{itemize}

>>>>

When studying how cilia paralysis and ventricular morphology affect intraventricular
flow and transport, we calculate the magnitude of a tangential force $F$ 
exerted by the cilia on a part $\mathcal{S}\subset\Gc$ of the ventricular surface as:
\begin{equation}
    F = \int_{\mathcal{S}}\sqrt{\btau\cdot\btau}\,\mathrm ds.
    \label{eq:tangential_force_calculation}
\end{equation}

@@

The CSF velocities are most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity magnitude drops from 27.7 \textmu m/s to 8.0 \textmu m/s.
In vivo, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic cilia has minimal
impact and results in maximum speeds of 27.7 \textmu m/s or 27.2 \textmu m/s, respectively.
Of the total tangential forces exerted by the cilia, ventral cilia in the middle ventricle
contribute with 34\%, while the dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). The magnitude of the force contributions are thus
more similar than the difference in their impact on the maximum velocity magnitudes.

Changes in the transport dynamics of Dendra2 are most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit changes in the dynamics of $\cbar$ (\Crefpanels{fig:fig5}{c}).
Compared to the baseline model results, we observe a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
In contrast, removal of telencephalic cilia does not alter the evolution of the
mean concentration $\cbar$ with time (\Crefpanels{fig:fig5}{c}).
Finally, without ventral cilia in the diencephalic ventricle,
more solute spreads towards the anterior ventricle (ROI 5) and
less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} The total tangential force applied with the tangential traction
    boundary condition in the three separate ciliated regions, as well as the total
    (the sum of the three regions), calculated with \Cref{eq:tangential_force_calculation}. 
    \textbf{c.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

\subsection{Modification of ventricles geometry impacts protein distribution}
Variations in ventricular morphology impact CSF flow patterns.
To assess how ventricular morphologies influence solute distribution,
we consider four alternative brain ventricles meshes:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Crefpanels{fig:fig6}{a}). In these
four modified geometries the volume is reduced by 3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively,
comapred to the original geometry.
We simulate CSF flow in the modified geometries with the baseline flow model
and use the resulting velocity field to simulate transport of Dendra2
(\Crefpanels{fig:fig6}{b}).

Constricting the connection between the anterior and middle ventricle leads to
slower transport to the anterior ventricle (ROI 5),
and a slightly higher mean concentration in ROIs 2--4 and 6, compared to the original geometry
(\Crefpanels{fig:fig6}{c}, shrunk fore-mid scenario). This is notwithstanding a
9.5\% reduction in ventral cilia force magnitude
(\Crefpanels{fig:fig6}{d}), which are the cilia that propel CSF and solute towards ROI 6.
On the other hand, constricting the duct connecting the middle and the posterior ventricle
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5)
(\Crefpanels{fig:fig6}{c}, shrunk mid-hind scenario).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a
shrunk middle ventricle.
Here also the change in transport to ROI 6 is counter to expectation when comparing with
ventral cilia forces, with transport reducing in spite of a 6.6\% increase in the
ventral cilia forces.
Additionally, for the shrunk middle configuration, the transport upstream of CSF flow towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.
For the mesh where all three ventricles are modified,
we observe higher retention of solute in the middle ventricle (ROIs 2--4),
while both the anterior and the posterior ventricles (ROIs 5 and 6) have lower mean concentrations.
Altogether, our results show that the geometry and width of the ventricular ducts matter
for solute transport, with notable differences in transport from relatively small
geometry modifications. This may explain why we observed larger differences
in the cilia-deficient computational model compared to the experimental data in
\Cref{fig:fig4}, as the model and the fish have different duct diameters.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure6_compare_modified_geometries.png}
    \caption{\textbf{a.} The original computational mesh of the ventricular geometry.
    \textbf{b.} The four modified versions of the computational geometry. Compared
    to the original mesh, the volumes of the deformed meshes are reduced by 3.9\%, 4.1\%, 1.4\% 
    and 9.4\%.
    \textbf{c.} Mean concentration $\cbar(t)$ simulated with the original and modified geometries
    in the 6 ROIs.
    \textbf{d.} A comparison of the tangential cilia force in the three cilia regions
    and the total force applied, calculated with \Cref{eq:tangential_force_calculation}.
    In each region, the force $F_{\mathrm{original}}$ 
    applied on the original mesh is the baseline of 100\%,
    to which the forces $F_i$ of the other mesh versions are compared, where $i$ denotes
    the cases fore-shrunk, middle shrunk, mid-hind shrunk, and the combination of the three.
    The bars for the four shrunk scenarios thus
    display $F_i/F_{\mathrm{original}}\cdot 100\%$.}
    \label{fig:fig6}
\end{figure}

@@

\subsection{Limitations and further work}
The model presented has certain limitations. Future work could consider improving:
\begin{itemize}
    \item quantify cilia force parameter to avoid tuning
    \item spatial heterogeneity of the cilia, how this affects flow and transport
    \item time-varying cilia?
    \item improved advection-diffusion equation influx/outflux BCs, or instead
    modeling deformation of a closed ventricular system. 
    and investigating how the arising flow compares with pulsatile flow observed in experiments
    \item FSI: CSF flow coupled to 1D/2D/3D cilia structures.
    \item any comments on experiments?
\end{itemize}
